---
title: "Analysis and Prediction of Shared Bicycle Usage at the University of Toronto St. George Campus"
subtitle: "Main Finding TBD"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "Students and staff at the University of Toronto's St. George campus often face difficulties finding a bike-sharing parking spot when arriving at campus, and finding a bike when leaving, which highlights the need for effective prediction of future bike-sharing demand to support campus commuting infrastructure. This study uses Bayesian Poisson regression to predict the usage of bike-sharing stations at 27 locations on campus for specific years, months, and  four-hour intervals of the day. The results show that bike-sharing usage on campus has been increasing annually, with significant seasonal variations and peak usage between 8 am and 8 pm."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
set.seed(912)
# List of necessary packages
packages <- c("tidyverse", "dplyr", "arrow", "sf", "ggplot2", "ggspatial", "RColorBrewer", "knitr")
# Install missing packages
missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
install.packages(missing_packages)
# Load packages
lapply(packages, library, character.only = TRUE)

#### Read data ####
start <- read_parquet(here::here("data/02-analysis_data/start.parquet"))
stop <- read_parquet(here::here("data/02-analysis_data/stop.parquet"))
```

# Introduction
Bike-sharing services have undoubtedly improved commuting efficiency, especially for short trips within a campus. For example, a study on bike-sharing in Xiamen, China, highlighted how data-driven analyses can inform policymakers about system performance, thereby guiding investments and policy decisions to enhance urban mobility[https://arxiv.org/abs/2401.03987?utm_source=chatgpt.com]. However, in the Toronto region, despite a study by the University of Toronto's School of Cities showing that ridership increased from about 665,000 trips in 2015 to over 4.5 million in 2022[https://schoolofcities.github.io/bike-share-toronto/growth], there has been little research focused on small-scale areas, such as universities, within Toronto. Furthermore, a study from Hangzhou, China, confirmed the significant differences in traveling characteristics between cities and university campuses[https://file.techscience.com/uploads/attached/file/20200916/20200916072445_75552.pdf?utm_source=chatgpt.com]. Therefore, this paper takes the University of Toronto as a case study to analyze bike-sharing usage within the campus.

The core of this study is to extract a sample of 27 bike-sharing stations within the University of Toronto campus from all bike-sharing usage data between January 1, 2017, and September 30, 2024. The total usage of bike-sharing at each station is calculated for each 4-hour interval. A Bayesian Poisson regression model is then employed to predict the usage based on the station, year, month, date, and the specific 4-hour interval of the day.

This study found that although the usage at all stations within the University of Toronto campus has increased rapidly each year, there are significant seasonal differences. During winter, usage at all stations, regardless of the year, remains close to zero. Additionally, from January to September each year, the usage shows an upward trend, which then gradually declines over the following months. There are also significant differences in usage across different times of the day. The period from 8 am to 8 pm accounts for an average of 79.8% of daily usage, while the period from midnight to 8 am accounts for only 6.8%. Notably, the number of bikes checked out and parked at the 27 stations is generally balanced; however, there are four stations where there is a substantial discrepancy between the number of check-outs and returns.

Compared to the general statistics on bike-sharing usage in the Greater Toronto Area, these findings offer more detailed recommendations specifically for transportation planning and policy adjustments within the University of Toronto campus. They also provide guidance on how to better allocate, deploy, or store shared bikes.

Structure: TBD

## Estimand
This study aims to estimate the usage of a specific bike-sharing station within the University of Toronto campus during a particular 4-hour interval. By converting time into year, month, day, and the specific 4-hour interval of the day, it accounts for the overall trend, seasonal effects, and hourly variation in station usage. The core objective is to explore the temporal changes in the usage of 27 bike-sharing stations within the campus, thereby providing policymakers with recommendations for bike allocation to improve commuting efficiency on campus.

# Data {#sec-data}

## Overview
The dataset used in this study comes from opendatatoronto [opendatatoronto], uploaded by Toronto Parking Authority and collected by Bike Share Toronto. It records every bike-sharing usage in the Toronto area from 2015 to September 30, 2024, with a total of 28,017,329 records. The variables included in the data differ across years, but they all contain the following variables: Trip ID, Trip Duration, Trip Start Station ID, Trip Start Time, Trip Start Station Location, Trip End Station ID, Trip End Time, Trip End Station Location, Bike ID, and User Type.

This study follows the workflow of Telling Stories with Data [Telling Stories with Data], using its initial folder structure and part of its code. Data downloading, cleaning, modeling, and visualization were carried out using R [R]. The following R libraries were also used alongside R:

TBD

## Measurement
	
To predict the relationship between bike-sharing usage and time within the University of Toronto's St. George campus, this study requires a time series to describe the changes in usage at different stations over time. Therefore, the dataset published by opendatatoronto [opendatatoronto], which records every instance of bike-sharing usage in the Toronto area since 2015, is ideal for this study. Additionally, this data is ideal because of its reliability—due to the commercial nature of bike-sharing, the specific time and location of each bike's use and return are accurately recorded. However, the raw data cannot be used directly in this project; sophisticated data cleaning is required, with specific steps and reasons as follows:

First, data from 2015-2016 was excluded because the collection and recording methods for those years differ significantly from later years and do not include the time and station variables required for this study. Next, data for stations located within the University of Toronto's St. George campus was extracted from all remaining samples, and the usage of each station was calculated for every four-hour interval. Finally, the data format was standardized for subsequent analysis. The cleaned data is in the format of @tbl-cleaned_data.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_data
#| tbl-cap: "Samples of the dataset used for analysis"
start |>
  tail() |>
  kable(booktabs = TRUE, align = "c")
```



## Variables

This study focuses on the following variables:

- `count`: The dependent variable of the study, a non-negative integer. It describes the total usage of bike-sharing at a particular station during a specific 4-hour interval.
- `time`: An independent variable representing the time interval. For example, "2024-09-29 12:00:00" represents the time interval from 12 pm to 4 pm on September 29, 2024, in a 24-hour format. The earliest `time` is "2017-01-01 00:00:00," and the latest is "2024-09-30 24:00:00."
- `station_name`: An independent variable representing the unique name of one of the 27 stations within the University of Toronto's St. George campus.


```{r}
#| label: fig-daily_usage
#| fig-cap: Daily usage from January 2017 to September 2024
#| echo: false
#| message: false
start_asdate <- start 
start_asdate$time <- as.Date(start_asdate$time)
daily_counts <- start_asdate %>%
  group_by(time) %>%
  summarise(total_count = sum(count))
ggplot(data = daily_counts, aes(x = time, y = total_count)) +
  geom_point() +
  labs(x = "Time", y = "Count") +
  theme_minimal()+  
  geom_smooth(method = "lm")  
```










<!-- # Model -->

<!-- The goal of our modelling strategy is twofold. Firstly,... -->

<!-- Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details]. -->

<!-- ## Model set-up -->

<!-- Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.   -->

<!-- \begin{align}  -->
<!-- y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\ -->
<!-- \mu_i &= \alpha + \beta_i + \gamma_i\\ -->
<!-- \alpha &\sim \mbox{Normal}(0, 2.5) \\ -->
<!-- \beta &\sim \mbox{Normal}(0, 2.5) \\ -->
<!-- \gamma &\sim \mbox{Normal}(0, 2.5) \\ -->
<!-- \sigma &\sim \mbox{Exponential}(1) -->
<!-- \end{align} -->

<!-- We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`. -->


<!-- ### Model justification -->

<!-- We expect a positive relationship between the size of the wings and time spent aloft. In particular... -->

<!-- We can use maths by including latex between dollar signs, for instance $\theta$. -->


<!-- # Results -->

<!-- Our results are summarized in @tbl-modelresults. -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| eval: true -->
<!-- #| warning: false -->
<!-- #| message: false -->

<!-- library(rstanarm) -->

<!-- first_model <- -->
<!--   readRDS(file = here::here("models/first_model.rds")) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| eval: true -->
<!-- #| label: tbl-modelresults -->
<!-- #| tbl-cap: "Explanatory models of flight time based on wing width and wing length" -->
<!-- #| warning: false -->

<!-- modelsummary::modelsummary( -->
<!--   list( -->
<!--     "First model" = first_model -->
<!--   ), -->
<!--   statistic = "mad", -->
<!--   fmt = 2 -->
<!-- ) -->
<!-- ``` -->




<!-- # Discussion -->

<!-- ## First discussion point {#sec-first-point} -->

<!-- If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.  -->

<!-- ## Second discussion point -->

<!-- Please don't use these as sub-heading labels - change them to be what your point actually is. -->

<!-- ## Third discussion point -->

<!-- ## Weaknesses and next steps -->

<!-- Weaknesses and next steps should also be included. -->

<!-- \newpage -->

<!-- \appendix -->

<!-- # Appendix {-} -->


<!-- # Additional data details -->

<!-- # Model details {#sec-model-details} -->

<!-- ## Posterior predictive check -->

<!-- In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows... -->

<!-- In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...  -->

<!-- ```{r} -->
<!-- #| eval: true -->
<!-- #| echo: false -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- #| label: fig-ppcheckandposteriorvsprior -->
<!-- #| layout-ncol: 2 -->
<!-- #| fig-cap: "Examining how the model fits, and is affected by, the data" -->
<!-- #| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"] -->

<!-- pp_check(first_model) + -->
<!--   theme_classic() + -->
<!--   theme(legend.position = "bottom") -->

<!-- posterior_vs_prior(first_model) + -->
<!--   theme_minimal() + -->
<!--   scale_color_brewer(palette = "Set1") + -->
<!--   theme(legend.position = "bottom") + -->
<!--   coord_flip() -->
<!-- ``` -->

<!-- ## Diagnostics -->

<!-- @fig-stanareyouokay-1 is a trace plot. It shows... This suggests... -->

<!-- @fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests... -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| eval: true -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- #| label: fig-stanareyouokay -->
<!-- #| fig-cap: "Checking the convergence of the MCMC algorithm" -->
<!-- #| fig-subcap: ["Trace plot", "Rhat plot"] -->
<!-- #| layout-ncol: 2 -->

<!-- plot(first_model, "trace") -->

<!-- plot(first_model, "rhat") -->
<!-- ``` -->



<!-- \newpage -->


<!-- # References -->


